---
title: "Storms and their impact on the community"
author: "Paul Hickman"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: false
---

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This analysis investigated the economic cost and impact on public health of storms and other severe weather events using the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.  The data was aggregated by event type for each of the main data points, property & crop damage, fatalities and injuries, before plotting the top 10 events for each.

## Data Processing

The data analysed is in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. The file is downloaded from the course web site and loaded for processing.

Further information about the file's structure can be found at the following locations:

- National Weather Service Storm Data Documentation
- https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
- National Climatic Data Center Storm Events FAQ
- https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

```{r cache=TRUE}
fname = "repdata-data-StormData.csv.bz2"
if(!file.exists(fname)){
  print("Downloading data...")
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", fname, method = "curl")
}
data = read.table(fname, sep=",", header = TRUE)
```

There is a mix of upper and lower case event types.  Make them all upper case so they will be aggregated together.

```{r}
# consistent case for event types
data$EVTYPE = toupper(data$EVTYPE)
```

Population health is defined as fatalities and injuries (per event type).  Fatalitiies and injuries are aggregated per each event type.

```{r}
# fatalities
fByType <- aggregate(FATALITIES ~ EVTYPE , data, sum, na.rm = TRUE)
fByType <- fByType[order(fByType$FATALITIES, decreasing = TRUE), ]
fByType <- fByType[1:10,] # interest in top 10

# injuries
iByType <- aggregate(INJURIES ~ EVTYPE , data, sum, na.rm = TRUE)
iByType <- iByType[order(iByType$INJURIES, decreasing = TRUE), ]
iByType <- iByType[1:10,] # interest in top 10
```

Economic consequences is defined as property and crop damage (per event type).

PROPDMGEXP and CROPDMGEXP indicate the units for PROPDMG and CROPDMG.

```{r}
# $ PROPDMGEXP: Factor w/ 19 levels
unique(data$PROPDMGEXP)
# $ CROPDMGEXP: Factor w/ 9 levels
unique(data$CROPDMGEXP)
```

Given the magnitude (EXP) column, calculate the total cost impact.

```{r}
library(car)
data$PROPDMG_TOT <- data$PROPDMG*as.numeric(recode(data$PROPDMGEXP, "'0'=10;'1'=10;'2'=10;'3'=10;'4'=10;'5'=10;'6'=10;'7'=10;'8'=10;'B'=10^9;'h'=10^2;'H'=10^2;'K'=10^3;'m'=10^6;'M'=10^6;'-'=0;'?'=0;'+'=1", as.factor.result = FALSE))

data$CROPDMG_TOT <- data$CROPDMG*as.numeric(recode(data$CROPDMGEXP, "'0'=10;'2'=10;'B'=10^9;'k'=10^3;'K'=10^3;'m'=10^6;'M'=10^6;''=0;'?'=0", as.factor.result = FALSE))
```

Property and crop damage are aggregated per each event type.

```{r}
# property damage
pByType <- aggregate(PROPDMG_TOT ~ EVTYPE , data, sum, na.rm = TRUE)
pByType <- pByType[order(pByType$PROPDMG_TOT, decreasing = TRUE), ]
pByType <- pByType[1:10,] # interest in top 10

# crop damage
cByType <- aggregate(CROPDMG_TOT ~ EVTYPE , data, sum, na.rm = TRUE)
cByType <- cByType[order(cByType$CROPDMG_TOT, decreasing = TRUE), ]
cByType <- cByType[1:10,] # interest in top 10
```

## Results

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The impact of tornadoes on public health is by far the greatest of all weather event types.

```{r}
# plots appear next to each other, shrink the text to fit the event labels and make some room at the bottom
par(mfrow = c(1,2), mar = c(8,4,4,2), cex.axis = 0.7, oma = c(0, 0, 2, 0))

barplot(fByType$FATALITIES, names.arg = fByType$EVTYPE, las = 3, col = "light green", main = "Fatalities", ylab = "Fatalities")

barplot(iByType$INJURIES, names.arg = iByType$EVTYPE, las = 3, col = "light green", main = "Injuries", ylab = "Injuries")

title("Top 10 Weather Events (1950-2011)", outer = TRUE)
```

Across the United States, which types of events have the greatest economic consequences?

Floods have had the greatest economic impact on property and drought has had the greatest economic impact on crops.

```{r}
# plots appear next to each other, shrink the text to fit the event labels and make some room at the bottom
par(mfrow = c(1,2), mar = c(8,4,4,2), cex.axis = 0.7, oma = c(0, 0, 2, 0))

barplot(pByType$PROPDMG_TOT/10^9, names.arg = pByType$EVTYPE, las = 3, col = "light green", main = "Property Damage", ylab = "Cost USD (billion)")

barplot(cByType$CROPDMG_TOT/10^9, names.arg = cByType$EVTYPE, las = 3, col = "light green", main = "Crop Damage", ylab = "Cost USD (billion)")

title("Top 10 Weather Events (1950-2011)", outer = TRUE)
```

### Conclusion

Over the 60 years (1950-2011) that data was collected, Tornadoes have had the greatest impact to public health (injuries and fatalities), floods have inflicted the greatest property damage and droughts are responsible for the greatest crop damage.
